---
title: "WBG Niger Climate-Resilient inter-urban Connectivity Project"
subtitle: "GCA Research for Impact"
date: 03-28-2023
date-format: long
format: 
  revealjs:
    logo: "R/gca_logo_blue_rgb.png"
    footer: "Global Center on Adaptation"
    fontcolor: "#0B5FBA"
    fontsize: 16px
    mainfont: "Roboto"
    echo: false
    slide-number: true
    embed-resources: true
---

::: {style="font-size:150%"}
## Data

-   **Population**

    -   Subnational (ADM1-3) population
    -   Settlements and density (GRID3)

-   **Roads & facilities (e.g., markets)**

    -   OpenStreetMaps

-   **Food security**

    -   Cadre Harmonisé (CH) & Integrated Phase Classification (IPC)
    -   Niger - Enquête Harmonisée sur le Conditions de Vie des Ménages 2018-2019 (World Bank)

-   **Climate/weather and Land use**

-   *Local trade, prices, production, etc. ? (edge weights)*
:::

## Preliminary data layering

```{r}
library(tidyverse)
library(ggplot2)
library(tmap)
library(mapview)

load("data/20230328_presentation.RData")
```

::: panel-tabset
## Population

```{r}
tmap_mode("view")

tm_shape(pop_sf) +
  tm_polygons("t_tl", palette = "Blues", n = 4, alpha = .5, title = "Population", group = "Population") +
  tm_shape(adm03) +
  tm_borders() +
  tm_scale_bar() +
  tm_minimap()
```

## Food security

```{r}
tmap_mode("view")

tm_shape(ipc_2020_adm2) +  
 tm_polygons("median", palette = "YlOrRd",  alpha = .5, id = "adm2_name", title = "Median % of pop. in IPC Phase 3-5", group = "Food Security") +
tm_shape(adm03) +
  tm_borders() +
tm_scale_bar() +
tm_minimap()
```

## Road network

```{r}
tmap_mode("view")

tm_shape(lines_ner) +
  tm_lines("edge_betweenness", palette = "YlOrRd", lwd = 3, 
           group = "Roads") +
tm_shape(lines_ner) +
  tm_lines("edge_betweenness_dist", palette = "YlOrRd", lwd = 3, 
           group = "Roads (weighted)") +
tm_shape(lines_ner) +
  tm_lines("highway", lwd = 3, group = "Type") +
tm_shape(nodes_ner) +
  tm_bubbles(col = "betweenness", palette = "YlOrRd", alpha = .5, scale = .1,
             group = "Roads") +
tm_shape(nodes_ner) +
  tm_bubbles(col = "betweenness_dist", palette = "YlOrRd", alpha = .5, scale = .1,
             group = "Roads (weighted)") +
tm_shape(adm03) +
  tm_borders() +
tm_scale_bar() +
tm_minimap()
```
:::

## Food security trends

```{r}
data_trends |> 
  filter(exercise_year > 2017) |> 
ggplot(aes(x = exercise_year)) +
  geom_ribbon(aes(ymin = min, ymax = max), fill = "honeydew2") +
  geom_line(aes(y = median), colour = "springgreen3") +
  theme_minimal() +
  theme(plot.title = element_text(face="bold")) +
  facet_wrap(~adm1_name, nrow = 3) +
  labs(title = "Median, Max. and Min. % of population in Phase 3-5 by region, 2018-2022",
       caption = "source: Integrated Food Security Phase Classification (IPC)", 
       x = "Year", y = "% of population per ADM2")
```

## Next steps

::: panel-tabset
## Bivariate

![[Nelson et al. (2019)](https://doi.org/10.1177/0361198118822272)](outputs/external_fig/edges_sierra_leone.png){width="400"}

## FS/road network

![[Espinet et Rozenberg (2018)](https://doi.org/10.1177/0361198118794043)](outputs/external_fig/criticality_mozambique.png){width="800"}

## FS framework

![[Marivoet et al. (2019)](https://www.sciencedirect.com/science/article/pii/S0305750X19300750)](outputs/external_fig/fs_typo.png){width="600"}
:::
