---
title: "Enabling/Constraining factors: Access"
format: html
editor: visual
---

```{r grappe}

load(here('data','grappes.RData'))

```

# Trade

**source**: EHCVM 2018-2019

```{r}

# import survey data
ehcvm_conso_ner2018 <- read_dta(here('data_raw', 'ner_2018_ehcvm', 'ehcvm_conso_ner2018.dta'))
ehcvm_welfare_ner2018 <- read_dta(here('data_raw', 'ner_2018_ehcvm', 'ehcvm_welfare_ner2018.dta'))

# prepare dataset
ehcvm_conso_agg <- ehcvm_conso_ner2018 %>% 
  select(hhid, vague, grappe, menage, codpr, depan) %>% 
  mutate(depan_grains = sum( if_else(codpr %in% 1:11, depan, 0) ), 
         depan_imp_rice = sum( if_else(codpr %in% 3:4, depan, 0) ),
         sh_imp_rice = 100 * depan_imp_rice / depan_grains,
         .by = hhid)

# create var sh_imp_rice by grappe
trade_by_grappe <- summarize(ehcvm_conso_agg, 
                             sh_imp_rice = mean(sh_imp_rice, na.rm = T), .by = grappe)

# merge
trade <- merge(trade_by_grappe, grappe_sf, by = 'grappe')

# save
save(trade, file = here('data', 'trade.RData'))
st_write(trade, here('layers', 'trade.gpkg'), delete_layer = TRUE)

```

# Food prices

**source**: WFP

```{r food_prices}

# load wfp data
set_rhdx_config(hdx_site = "prod") # from the rhdx package
get_rhdx_config()

search_datasets("Niger - Food Prices", rows = 5)

ner_food_prices <- pull_dataset("wfp-food-prices-for-niger") %>%
  get_resource(1) %>%
  read_resource()

# prepare the dataset
food_prices <- ner_food_prices %>% 
  mutate(year = floor_date(date, "year")) %>% 
  filter(year >= '2010-01-01' & year < '2023-01-01') %>% 
  mutate(season = case_when(
    month(date) %in% c(12, 01:04) ~ 'dry',
    month(date) %in% c(05:07) ~ 'wet',
    .default = "other"
    )) # based on IPC periods

market_coverage <- food_prices %>% summarize(max = max(date), min = min(date), .by = "market")

food_prices <- food_prices %>% 
  filter(!market %in% c('Chetimari', 'Falwel', 'Dan-Goulbi', 'Chadakori', 
                        'Kanembakache', 'Maidjirgui', 'Jataka', 'Bonfeba', 
                        'Soubdou', 'Sassoumbroum', 'Guidimouni')) %>% # filter out markets with coverage <= 2010
  filter(priceflag == 'actual') %>% 
  filter(!commodity == 'Beans (niebe)') # coverage stops im 2014

# create food_prices var by market
food_price_by_market <- food_prices %>% 
  select(-c(category, unit, currency, usdprice, year)) %>% 
  filter(pricetype == 'Retail' & commodity %in% c('Maize', 'Millet', 'Sorghum')) %>%
  group_by(month = floor_date(date, "month"), commodity) %>% 
  mutate(mean_month = mean(price),
         local_price_diff = 100 * price / mean_month) %>% 
  ungroup() %>% 
  summarize(food_prices_avg = mean(local_price_diff, na.rm = T), 
            food_prices_med = median(local_price_diff, na.rm = T),
            diff = food_prices_avg - food_prices_med,
            .by = c(market, longitude, latitude)) %>% 
  arrange(desc(diff)) %>% 
  st_as_sf(coords = c("longitude", "latitude")) %>%
  st_set_crs(4326)

# join with grappe
food_price_by_grappe <- st_join(grappe_sf, 
                                    food_price_by_market, 
                                    join = nngeo::st_nn, k = 3, progress = FALSE) %>% 
  st_drop_geometry() %>% 
  summarize(food_prices_avg = mean(food_prices_avg),
            food_prices_med = mean(food_prices_med),
            .by = grappe) %>% 
  arrange(grappe)

# merge
food_price <- food_price_by_grappe %>% 
  merge(grappe_sf, by = 'grappe') %>% 
  st_as_sf()
  
# save
save(food_price, file = here('data', 'food_price.RData'))
st_write(food_price, here('layers', 'food_price.gpkg'), delete_layer = TRUE)

```

# Assets

**source**: EHCVM 2018-2019

```{r assets}

# load survey data
survey_welfare <- read_dta(here('data_raw', 'ner_2018_ehcvm', 'ehcvm_welfare_ner2018.dta'))
s17_me_agric <- read_dta(here('data_raw', 'ner_2018_ehcvm', 's17_me_ner2018.dta'))

s17_data <- s17_me_agric %>% 
  select(grappe, menage, 
         espece_code1 = s17q01,
         espece_code2 = s17q02,
         qte_livestock = s17q05) %>% 
  mutate(espece_code1 = as_factor(espece_code1),
         espece_code2 = as_factor(espece_code2))

livestock <- left_join(survey_welfare %>%
                         select(grappe, menage, hhid),
                       s17_data,
                       by = c("grappe", "menage")) %>%
  mutate(
    TLU = case_when(
      espece_code2 == 'Bovins (boeufs)' ~ .5,
      espece_code2 == 'Ovins (Moutons)' ~ .1,
      espece_code2 == 'Caprins (Chèvres)' ~ .1,
      espece_code2 == 'Camelins (Chameaux)' ~ .7,
      espece_code2 == 'Equins (Chevaux)' ~ .5,
      espece_code2 == 'Asins (Anes)' ~ .6,
      espece_code2 == 'Porcins (Cochons)' ~ .2,
      espece_code2 == 'Lapins' ~ .02,
      espece_code2 == 'Poules / poulets' ~ .01,
      espece_code2 == 'Pintades' ~ .01,
      espece_code2 == 'Autres volailles' ~ .01
    )
  ) %>% 
  mutate(tlu_grappe = sum(qte_livestock * TLU, na.rm = T), .by = grappe) %>% 
  mutate(tlu_total = sum(qte_livestock * TLU, na.rm = T))
         
# create var sh_livestock by grappe
livestock_by_grappe <- livestock %>%
  reframe(sh_livestock = tlu_grappe / tlu_total, .by = grappe) %>% 
  unique()
  
# merge
livestock <- merge(livestock_by_grappe, grappe_sf, by = 'grappe')

# save
save(livestock, livestock_by_grappe, file = here('data', 'livestock.RData'))
st_write(livestock, here('layers', 'livestock.gpkg'), delete_layer = TRUE)

```

| **Animal category**    | **TLU (Ahmed and Mesfin, 2017)** | **TLU (Rout, Kumar, and Behera, 2019)** |
|:-----------------------|:---------------------------------|:----------------------------------------|
| Calf                   | 0.25                             |                                         |
| Donkey (young)         | 0.35                             |                                         |
| Weaned Calf            | 0.34                             |                                         |
| Camel                  | 1.25                             | 0.70                                    |
| Heifer                 | 0.75                             |                                         |
| Sheep and goat (adult) | 0.13                             | 0.10                                    |
| Cow and ox             | 1.00                             | 0.50                                    |
| Sheep and goat (young) | 0.06                             |                                         |
| Horse                  | 1.10                             | 0.50                                    |
| Chicken                | 0.013                            | 0.01                                    |
| Donkey (adult)         | 0.70                             | 0.60                                    |
| Rabbit                 |                                  | 0.02                                    |
| Pig                    |                                  | 0.20                                    |

: Source: (1) [The impact of agricultural cooperatives membership on the wellbeing of smallholder farmers: empirical evidence from eastern Ethiopia \| Agricultural and Food Economics \| Full Text (springeropen.com)](http://127.0.0.1:40547/#0){style="text-decoration: underline; color: rgb(137, 89, 168) !important;"}; (2) [Goat production challenges to food security. \| Goat production and supply chain management in the tropics (cabidigitallibrary.org)](http://127.0.0.1:40547/#0){style="text-decoration: underline; color: rgb(137, 89, 168) !important;"}

# Income

**source**: EHCVM 2018-2019

```{r}

# load survey data
s20_me_ner2018 <- read_dta(here('data_raw', 'ner_2018_ehcvm', 's20_me_ner2018.dta'))
survey_welfare <- read_dta(here('data_raw', 'ner_2018_ehcvm', 'ehcvm_welfare_ner2018.dta'))

# prepare dataset
s20_data <- s20_me_ner2018 %>%
  select(grappe, menage, relative_poverty = s20q02) %>% 
  left_join(survey_welfare %>% select(grappe, menage, hhid), by = c('grappe', 'menage'))

# create income var by grappe
income_by_grappe <- s20_data %>% 
  group_by(grappe, relative_poverty) %>% 
  summarize(n = n()) %>% 
  mutate(income = 100 * n / sum(n)) %>% 
  filter(relative_poverty == 4) %>% 
  ungroup()
            
# merge
income <- left_join(grappe_sf, income_by_grappe, by = 'grappe') %>% 
  select(-c(relative_poverty, n)) %>% 
  mutate(income = replace_na(income, 0))

# save
save(income, file = here('data', 'income.RData'))
st_write(income, here('layers', 'income.gpkg'), delete_layer = TRUE)

```
